---
title: "Who is most exposed to generative AI? Occupations and demographics in the Australian workforce"
subtitle: "Assignment 4 ETC5512"
author: "Who are you?? Add your name here"
format: html
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

1.  **Main Research Question**

    How do demographic characteristics influence exposure to generative AI across occupations in Australia?

    **Sub-Questions**

    -   Which occupations exhibit the highest and lowest levels of generative AI exposure scores?

    -   Which demographic groups (age and sex) are most concentrated in occupations with high AI-exposure?

    -   To what extent are highly AI-exposed occupations characterised by AI augmentation versus AI automation?

    **Motivation**

    Generative AI has rapidly become one of the most significant technological developments shaping the world and particularly modern workplaces. More than 50% of organisations in Australia have reportedly integrated AI into their operations (CSIRO, 2025), and the adoption of AI is expected to continue rising with significant impact across different occupations and industries. Jobs and Skills Australia (2025) identifies generative AI as a major force likely to influence the Australian labour market, with effects depending on occupational tasks, workplace adoption, and workforce adaptation.

    Understanding AI exposure is becoming increasingly important as the impacts of AI use are likely to be unevenly distributed across the workforce. Jobs and Skills Australia (2025) suggests that generative AI is more likely to augment human work than replace it, where the greatest potential for automation is concentrated in more routine tasks. The report also highlighted substantial differences in AI exposure across occupations, indicating that some workers may experience greater changes to their work than others.

    This topic was chosen because it combines a highly relevant contemporary issue with publicly available Australian labour market data and demographic census data. By examining how AI exposure is distributed across occupations and demographic groups, the analysis aims to identify which occupations exhibit the most extreme levels of AI exposure, and whether these highly exposed occupations are characterised more by AI augmentation or AI automation. This contributes to a better understanding of how exposure to generative AI is distributed across the Australian workforce.

2.  **Data Sources**

    To investigate how demographic characteristics are associated with exposure to generative AI across occupations in Australia, two datasets will be combined:

    -   Australian Bureau of Statistics (ABS) Census Data

    -   Jobs and Skills Australia Generative AI Exposure Data

    The first dataset is the 2021 Australian Census G60: Occupation by Age by Sex Geopackage published by the Australian Bureau of Statistics (ABS). The data is from the 2021 population census which aimed to collect information from all people in Australia on census night in 2021. The dataset contains counts of employed persons by occupation, age group, sex, and geographic region, providing detailed demographic information about the Australian workforce.

    This dataset is appropriate for the analysis because it provides comprehensive demographic information on the Australian workforce that can be linked to occupation-level AI exposure measures using occupation codes or classifications. The data enables analysis of the age and sex composition of different occupational groups and the demographic distribution of workers across occupations with varying levels of AI exposure. Using Census data also reduces sampling error that may be present in survey-based datasets due to its near-complete population coverage.

    The data was released under the Creative Commons Attribution 4.0 International (CC BY 4.0) licence, allowing reuse with appropriate attribution to the ABS.

    Limitations of the dataset include that it reflects workforce characteristics at the time of the 2021 Census and may not fully represent current labour market conditions, as workplace composition and occupation classifications may have changed between 2021 to 2025. Furthermore, occupation categories are aggregated and may conceal variation within occupations.

    From an ethical and privacy perspective, the ABS applies confidentiality procedures before releasing Census data. The dataset contains aggregated counts only and does not include any personally identifiable information, resulting in minimal privacy concerns.

    The second dataset is the occupation-level Generative AI Exposure dataset published by Jobs and Skills Australia as part of their "*Our Gen AI Transition: Exposure, Adaptation and Dynamism*" study. The dataset is a model-based research dataset containing estimated measures of generative AI exposure for occupations, including indicators describing the extent to which AI may augment or automate occupational tasks within an occupation.

    This dataset is suitable for this analysis because it directly measures the potential exposure of occupations to generative AI technologies. Linking these exposure scores to Census occupation data enables the analysis to build on Jobs and Skills Australia's existing report and identify how AI exposure is distributed across different demographic groups within highly exposed occupations, and compare the relative importance of augmentation and automation in the Australian workforce.

    The dataset is publicly available through Jobs and Skills Australia and is released under the Australian Government's Creative Commons Attribution (CC BY) licence.

    A key limitation of this dataset is that the exposure measures are model-based estimates rather than real observed workplace outcomes. The scores indicate the potential exposure of generative AI on occupational tasks and should not be interpreted as true measures of AI adoption or job displacement in Australia. Additionally, the estimates may change over time as AI technologies and workplace practices continue to evolve.

    Ethically, care must be taken when interpreting AI exposure scores such that high exposure does not necessarily imply that jobs will be replaced, as many occupations are expected to experience task augmentation rather than full automation. Results should therefore be interpreted as indicators of occupational exposure to generative AI rather than predictions of employment outcomes and job replacement.

3.  **Data Download**

    The two datasets were directly downloaded from their respective sources and stored separately to preserve the original files.

    For the Australian Bureau of Statistics (ABS) Census Data:

    1.  Go to [Search Census GeoPackages](https://www.abs.gov.au/census/find-census-data/geopackages?release=2021&geography=AUST&table=G60&gda=GDA2020) on the Australian Bureau of Statistics webpage.
    2.  For **census date** select *2021*.
    3.  For **state or territory** select *All geographies*.
    4.  For **table or topic** select *Table*.
    5.  Select *G60 - Occupation by age by sex* from the drop-down list.
    6.  For **GDA type** select GDA2020.
    7.  Download the ZIP file for the G60 - Occupation by age by sex GeoPackage as `Geopackage_2021_G60_AUST_GDA2020.zip`
    8.  Unzip the file and move the folder into the `raw_data` directory.

    For the Jobs and Skills Australia Generative AI Exposure Data:

    1.  Go to [Our Gen AI Transition](https://www.jobsandskills.gov.au/studies/generative-artificial-intelligence-capacity-study#datatableanddownlo) on the Jobs and Skills Australia website.
    2.  Scroll down to the *Data table and downloads* section.
    3.  Click the **Download** button for the *JSA AI Study Interactive Tables* under the Download subsection.
    4.  Save the downloaded XLS data as `raw_exposure_data.xlsx` into the `raw_data` directory.

```{r}
# Load in required packages for reading in raw data
library(readxl)
library(sf)

# Read in the raw data from the raw_data directory
raw_exposure_data <- read_excel("raw_data/raw_exposure_data.xlsx", sheet = "Occupation", skip = 6)

geopath_2021_G60A <- here::here("raw_data/Geopackage_2021_G60_AUST_GDA2020/G60A_AUST_GDA2020.gpkg")

st_layers(geopath_2021_G60A)

raw_census_data <-read_sf(geopath_2021_G60A, layer = "G60A_STE_2021_AUST")
```

**Data Processing Steps**

1.  Explore the raw exposure dataset.

```{r}
# Load in tidyverse package
library(tidyverse)

summary(raw_exposure_data)
```

2.  Remove non-data rows.

```{r}
clean_exposure_data <- raw_exposure_data %>%
  slice(1:(n() - 2))
```

3.  Aggregate occupations into major groups.

```{r}
clean_exposure_data <- clean_exposure_data %>%
  mutate(major_group_code = substr(`ANZSCO unit code`, 1, 1),
         major_occupation_group = case_when(
           major_group_code == "1" ~ "Managers",
           major_group_code == "2" ~ "Professionals",
           major_group_code == "3" ~ "Technicians and Trades Workers",
           major_group_code == "4" ~ "Community and Personal Service Workers",
           major_group_code == "5" ~ "Clerical and Administrative Workers",
           major_group_code == "6" ~ "Sales Workers",
           major_group_code == "7" ~ "Machinery Operators and Drivers",
           major_group_code == "8" ~ "Labourers"
         ))
```

4.  Aggregate occupation-level AI exposure measures into mean values for each ANZSCO major occupation group and retain relevant variables only.

```{r}
clean_exposure_data <- clean_exposure_data %>%
  group_by(major_occupation_group) %>%
  summarise(ai_augmentation_score = mean(`Augmentation exposure score`, na.rm = TRUE),
            ai_automation_score = mean(`Automation exposure score`, na.rm = TRUE),
            hybridisation = mean(`Hybridisation potential (differential score; scaled  x 1000)`, na.rm = TRUE),
            specialisation = mean(`Specialisation potential (differential score; scaled x 1000)`, na.rm = TRUE),
            .groups = "drop") %>%
  select(major_occupation_group, ai_augmentation_score, ai_automation_score, hybridisation, specialisation)
```

5.  Explore the raw census dataset.

```{r}
summary(raw_census_data)
```

6.  Use string manipulation to split the demographics.

```{r}
names(raw_census_data) <- names(raw_census_data) %>%
  str_replace_all("85_?ov", "85_99")

clean_census_data <- raw_census_data %>%
  st_drop_geometry() %>%
  pivot_longer(cols = -c(STE_NAME_2021, STE_CODE_2021, AREA_ALBERS_SQKM), names_to = "variable", values_to = "count")

clean_census_data <- clean_census_data %>%
  extract(
    variable,
    into = c("sex", "age_group", "occupation"),
    regex = "^([MF])([0-9_]+)_(.*)$"
  ) %>%
  mutate(
    sex = case_when(
      sex == "M" ~ "Male",
      sex == "F" ~ "Female"
    ),
    age_group = case_when(
      age_group == "85_99" ~ "85+",
      TRUE ~ str_replace_all(age_group, "_", "-")
    )
  )
```

7.  Align occupation classification names between the ABS Census data and the JSA exposure dataset to enable merging.

```{r}
unique(clean_census_data$occupation)

clean_census_data <- clean_census_data %>%
  mutate(major_occupation_group = case_when(
    occupation == "CommunPersnlSvc_W" ~ "Community and Personal Service Workers",
    occupation == "ClericalAdminis_W" ~ "Clerical and Administrative Workers",
    occupation == "Sales_W" ~ "Sales Workers",
    occupation == "Mach_oper_drivers" ~ "Machinery Operators and Drivers",
    occupation %in% c("TechnicTrades_Wrs", "TechnicTrades_W") ~ "Technicians and Trades Workers",
    TRUE ~ occupation))
  
```

8.  Retain only ANZSCO major occupation groups that can be directly matched to the JSA exposure data (e.g. totals).

```{r}
unique(clean_census_data$major_occupation_group)

clean_census_data <- clean_census_data %>%
  filter(major_occupation_group %in% c(
    "Managers", 
    "Professionals", 
    "Technicians and Trades Workers", 
    "Community and Personal Service Workers", 
    "Clerical and Administrative Workers", 
    "Sales Workers", 
    "Machinery Operators and Drivers", 
    "Labourers"))

unique(clean_census_data$major_occupation_group)
```

9.  Keep relevant variables only.

```{r}
clean_census_data <- clean_census_data %>%
  rename(state = STE_NAME_2021) %>%
  select(-STE_CODE_2021, -AREA_ALBERS_SQKM, -occupation)
```

10. Merge the cleaned ABS Census and JSA datasets using the major occupation group classification.

```{r}
mismatch_check <- anti_join(clean_census_data, clean_exposure_data, by = "major_occupation_group")
mismatch_check

processed_data <- clean_census_data %>%
  left_join(clean_exposure_data, by = "major_occupation_group")

sum(is.na(processed_data$ai_augmentation_score))
sum(is.na(processed_data$ai_automation_score))
```

11. Save the processed data into the `processed_data` directory.

```{r}
write_rds(processed_data, "processed_data/merged_data.rds")
```

**Additional Files for Marker Review**

In addition to the main analysis report, the following files are provided for marker review.

-   README.txt: This file contains detailed documentation of the project, including data origin, files included, intended use, assumptions, limitations, and reproducibility instructions.

-   data_dictionary.csv: This file describes all variables in the final processed dataset, including variable names, data types, descriptions and derivation methods.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

**Motivation**

Generative AI has rapidly become one of the most significant technological advancements shaping modern workplaces. In Australia, AI adoption is increasing across industries, with Jobs and Skills Australia highlighting generative AI as a key force likely to influence how work is performed rather than simply replacing jobs.

However, AI exposure is not expected to be uniform across the workforce. Different occupations involve different task structures, meaning some roles are more likely to be affected by AI systems than others. At the same times, the demographic composition of occupations may shape which groups of workers are most represented in highly AI-exposed roles.

This blog explores how generative AI exposure varies across occupations in the Australian workforce and examines how demographic characteristics such as age and sex is distributed across occupations with differing levels of AI exposure. It also compares different two types of AI exposure: augmentation (where AI supports human work) and automation (where AI replaces tasks).

**Data**

This analysis combines two publicly available datasets.

The first is the Australian Bureau of Statistics (ABS) 2021 Census G60 Occupation by Age by Sex dataset. It provides counts of employed persons across Australian states, age groups, sex, and occupation categories. This dataset provides a detailed view of the demographic structure of the Australian workforce.

The second dataset is the Jobs and Skills Australia (JSA) Generative AI Exposure dataset. It contains occupation-level estimates of AI exposure, including measures of augmentation and automation potential. These scores are model-based estimates of how susceptible occupational tasks are to generative AI technologies.

The datasets were merged using ANZSCO major occupation group classifications to ensure comparability between occupational structures.

**Analysis**

```{r}
occupation_exposure <- processed_data %>%
  distinct(major_occupation_group,
           ai_augmentation_score,
           ai_automation_score) %>%
  mutate(ai_exposure_score = (ai_augmentation_score + ai_automation_score) / 2)

occupation_exposure %>%
  arrange(desc(ai_exposure_score))

ggplot(
  occupation_exposure,
  aes(
    x = reorder(
      major_occupation_group,
      ai_exposure_score
    ),
    y = ai_exposure_score
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "AI Exposure Score by Occupation",
    x = "Occupation Group",
    y = "AI Exposure Score"
  )
```

**Conclusions**

**References**

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
