---
title: "Who is most exposed to generative AI? Occupations and demographics in the Australian workforce"
subtitle: "Assignment 4 ETC5512"
author: "Who are you?? Add your name here"
format: html
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

1.  **Main Research Question**

    How are demographic characteristics associated with employment in occupations with different levels of generative AI exposure in Australia?

    **Sub-Questions**

    -   Which occupations exhibit the highest and lowest levels of generative AI exposure scores?

    -   Which demographic groups (age and sex) are most concentrated in occupations with high AI-exposure?

    -   To what extent are highly AI-exposed occupations characterised by AI augmentation versus AI automation?

    **Motivation**

    Generative AI has rapidly become one of the most significant technological developments shaping the world and particularly modern workplaces. More than 50% of organisations in Australia have reportedly integrated AI into their operations (CSIRO, 2025), and the adoption of AI is expected to continue rising with significant impact across different occupations and industries. Jobs and Skills Australia (2025) identifies generative AI as a major force likely to influence the Australian labour market, with effects depending on occupational tasks, workplace adoption, and workforce adaptation.

    Understanding AI exposure is becoming increasingly important as the impacts of AI use are likely to be unevenly distributed across the workforce. Jobs and Skills Australia (2025) suggests that generative AI is more likely to augment human work than replace it, where the greatest potential for automation is concentrated in more routine tasks. The report also highlighted substantial differences in AI exposure across occupations, indicating that some workers may experience greater changes to their work than others.

    This topic was chosen because it combines a highly relevant contemporary issue with publicly available Australian labour market data and demographic census data. By examining how AI exposure is distributed across occupations and demographic groups, the analysis aims to identify how AI exposure is distributed across occupations and examine differences in the relative importance of AI augmentation and automation across occupations. This contributes to a better understanding of how exposure to generative AI is distributed across the Australian workforce.

2.  **Data Sources**

    To investigate how demographic characteristics are associated with exposure to generative AI across occupations in Australia, two datasets will be combined:

    -   Australian Bureau of Statistics (ABS) Census Data

    -   Jobs and Skills Australia Generative AI Exposure Data

    The first dataset is the 2021 Australian Census G60: Occupation by Age by Sex Geopackage published by the Australian Bureau of Statistics (ABS). The data is from the 2021 population census which aimed to collect information from all people in Australia on census night in 2021. The dataset contains counts of employed persons by occupation, age group, sex, and geographic region, providing detailed demographic information about the Australian workforce.

    This dataset is appropriate for the analysis because it provides comprehensive demographic information on the Australian workforce that can be linked to occupation-level AI exposure measures using occupation codes or classifications. The data enables analysis of the age and sex composition of different occupational groups and the demographic distribution of workers across occupations with varying levels of AI exposure. Using Census data also reduces sampling error that may be present in survey-based datasets due to its near-complete population coverage.

    The data was released under the Creative Commons Attribution 4.0 International (CC BY 4.0) licence, allowing reuse with appropriate attribution to the ABS.

    Limitations of the dataset include that it reflects workforce characteristics at the time of the 2021 Census and may not fully represent current labour market conditions, as workplace composition and occupation classifications may have changed between 2021 to 2025. Furthermore, occupation categories are aggregated and may conceal variation within occupations.

    From an ethical and privacy perspective, the ABS applies confidentiality procedures before releasing Census data. The dataset contains aggregated counts only and does not include any personally identifiable information, resulting in minimal privacy concerns.

    The second dataset is the occupation-level Generative AI Exposure dataset published by Jobs and Skills Australia as part of their "*Our Gen AI Transition: Exposure, Adaptation and Dynamism*" study. The dataset is a model-based research dataset containing estimated measures of generative AI exposure for occupations, including indicators describing the extent to which AI may augment or automate occupational tasks within an occupation.

    This dataset is suitable for this analysis because it directly measures the potential exposure of occupations to generative AI technologies. Linking these exposure scores to Census occupation data enables the analysis to build on Jobs and Skills Australia's existing report and examine how demographic groups are distributed across occupations with different levels of AI exposure, and compare AI augmentation and automation scores across occupations in the Australian workforce.

    The dataset is publicly available through Jobs and Skills Australia and is released under the Australian Government's Creative Commons Attribution (CC BY) licence.

    A key limitation of this dataset is that the exposure measures are model-based estimates rather than real observed workplace outcomes. The scores indicate the potential exposure of generative AI on occupational tasks and should not be interpreted as true measures of AI adoption or job displacement in Australia. Additionally, the estimates may change over time as AI technologies and workplace practices continue to evolve.

    Ethically, care must be taken when interpreting AI exposure scores such that high exposure does not necessarily imply that jobs will be replaced, as many occupations are expected to experience task augmentation rather than full automation. Results should therefore be interpreted as indicators of occupational exposure to generative AI rather than predictions of employment outcomes and job replacement.

3.  **Data Download**

    The two datasets were directly downloaded from their respective sources and stored separately to preserve the original files.

    For the Australian Bureau of Statistics (ABS) Census Data:

    1.  Go to [Search Census GeoPackages](https://www.abs.gov.au/census/find-census-data/geopackages?release=2021&geography=AUST&table=G60&gda=GDA2020) on the Australian Bureau of Statistics webpage.
    2.  For **census date** select *2021*.
    3.  For **state or territory** select *All geographies*.
    4.  For **table or topic** select *Table*.
    5.  Select *G60 - Occupation by age by sex* from the drop-down list.
    6.  For **GDA type** select GDA2020.
    7.  Download the ZIP file for the G60 - Occupation by age by sex GeoPackage as `Geopackage_2021_G60_AUST_GDA2020.zip`
    8.  Unzip the file and move the folder into the `raw_data` directory.

    For the Jobs and Skills Australia Generative AI Exposure Data:

    1.  Go to [Our Gen AI Transition](https://www.jobsandskills.gov.au/studies/generative-artificial-intelligence-capacity-study#datatableanddownlo) on the Jobs and Skills Australia website.
    2.  Scroll down to the *Data table and downloads* section.
    3.  Click the **Download** button for the *JSA AI Study Interactive Tables* under the Download subsection.
    4.  Save the downloaded XLS data as `raw_exposure_data.xlsx` into the `raw_data` directory.

```{r}
# Load in required packages for reading in raw data
library(readxl)
library(sf)

# Read in the raw data from the raw_data directory
raw_exposure_data <- read_excel("raw_data/raw_exposure_data.xlsx", sheet = "Occupation", skip = 6)

geopath_2021_G60A <- here::here("raw_data/Geopackage_2021_G60_AUST_GDA2020/G60A_AUST_GDA2020.gpkg")

st_layers(geopath_2021_G60A)

raw_census_data <-read_sf(geopath_2021_G60A, layer = "G60A_STE_2021_AUST")
```

**Data Processing Steps**

1.  Explore the raw exposure dataset.

```{r}
# Load in tidyverse package
library(tidyverse)

summary(raw_exposure_data)
```

2.  Remove non-data rows.

```{r}
clean_exposure_data <- raw_exposure_data %>%
  slice(1:(n() - 2))
```

3.  Aggregate occupations into major groups.

```{r}
clean_exposure_data <- clean_exposure_data %>%
  mutate(major_group_code = substr(`ANZSCO unit code`, 1, 1),
         major_occupation_group = case_when(
           major_group_code == "1" ~ "Managers",
           major_group_code == "2" ~ "Professionals",
           major_group_code == "3" ~ "Technicians and Trades Workers",
           major_group_code == "4" ~ "Community and Personal Service Workers",
           major_group_code == "5" ~ "Clerical and Administrative Workers",
           major_group_code == "6" ~ "Sales Workers",
           major_group_code == "7" ~ "Machinery Operators and Drivers",
           major_group_code == "8" ~ "Labourers"
         ))
```

4.  Aggregate occupation-level AI exposure measures into mean values for each ANZSCO major occupation group and retain relevant variables only.

```{r}
clean_exposure_data <- clean_exposure_data %>%
  group_by(major_occupation_group) %>%
  summarise(ai_augmentation_score = mean(`Augmentation exposure score`, na.rm = TRUE),
            ai_automation_score = mean(`Automation exposure score`, na.rm = TRUE),
            hybridisation = mean(`Hybridisation potential (differential score; scaled  x 1000)`, na.rm = TRUE),
            specialisation = mean(`Specialisation potential (differential score; scaled x 1000)`, na.rm = TRUE),
            .groups = "drop") %>%
  select(major_occupation_group, ai_augmentation_score, ai_automation_score, hybridisation, specialisation)
```

5.  Explore the raw census dataset.

```{r}
summary(raw_census_data)
```

6.  Use string manipulation to split the demographics.

```{r}
names(raw_census_data) <- names(raw_census_data) %>%
  str_replace_all("85_?ov", "85_99")

clean_census_data <- raw_census_data %>%
  st_drop_geometry() %>%
  pivot_longer(cols = -c(STE_NAME_2021, STE_CODE_2021, AREA_ALBERS_SQKM), names_to = "variable", values_to = "count")

clean_census_data <- clean_census_data %>%
  extract(
    variable,
    into = c("sex", "age_group", "occupation"),
    regex = "^([MF])([0-9_]+)_(.*)$"
  ) %>%
  mutate(
    sex = case_when(
      sex == "M" ~ "Male",
      sex == "F" ~ "Female"
    ),
    age_group = case_when(
      age_group == "85_99" ~ "85+",
      TRUE ~ str_replace_all(age_group, "_", "-")
    )
  )
```

7.  Align occupation classification names between the ABS Census data and the JSA exposure dataset to enable merging.

```{r}
unique(clean_census_data$occupation)

clean_census_data <- clean_census_data %>%
  mutate(major_occupation_group = case_when(
    occupation == "CommunPersnlSvc_W" ~ "Community and Personal Service Workers",
    occupation == "ClericalAdminis_W" ~ "Clerical and Administrative Workers",
    occupation == "Sales_W" ~ "Sales Workers",
    occupation == "Mach_oper_drivers" ~ "Machinery Operators and Drivers",
    occupation %in% c("TechnicTrades_Wrs", "TechnicTrades_W") ~ "Technicians and Trades Workers",
    TRUE ~ occupation))
  
```

8.  Retain only ANZSCO major occupation groups that can be directly matched to the JSA exposure data (e.g. totals).

```{r}
unique(clean_census_data$major_occupation_group)

clean_census_data <- clean_census_data %>%
  filter(major_occupation_group %in% c(
    "Managers", 
    "Professionals", 
    "Technicians and Trades Workers", 
    "Community and Personal Service Workers", 
    "Clerical and Administrative Workers", 
    "Sales Workers", 
    "Machinery Operators and Drivers", 
    "Labourers"))

unique(clean_census_data$major_occupation_group)
```

9.  Keep relevant variables only.

```{r}
clean_census_data <- clean_census_data %>%
  rename(state = STE_NAME_2021) %>%
  select(-STE_CODE_2021, -AREA_ALBERS_SQKM, -occupation)
```

10. Merge the cleaned ABS Census and JSA datasets using the major occupation group classification.

```{r}
mismatch_check <- anti_join(clean_census_data, clean_exposure_data, by = "major_occupation_group")
mismatch_check

processed_data <- clean_census_data %>%
  left_join(clean_exposure_data, by = "major_occupation_group")

sum(is.na(processed_data$ai_augmentation_score))
sum(is.na(processed_data$ai_automation_score))
```

11. Derive AI balance from augmentation score minus automation score.

```{r}
processed_data <- processed_data %>%
  mutate(ai_balance = ai_augmentation_score - ai_automation_score)
```

12. Save the processed data into the `processed_data` directory.

```{r}
write_rds(processed_data, "processed_data/merged_data.rds")
```

13. Check distribution of workers by age group.

```{r}
age_distribution <- processed_data %>%
  group_by(age_group) %>%
  summarise(total = sum(count), .groups = "drop")

ggplot(age_distribution, aes(x = age_group, y = total)) + geom_col() +
  labs(
    title = "Distribution of Workers by Age Group",
    x = "Age Group",
    y = "Number of Workers"
  )
```

Checking the distribution of rows in each age group shows that the workforce is unevenly distributed across age groups. The majority of workers are concentrated in the middle age categories (25-54 years), while younger workers (15-24 years) and older workers (65+ years) represent a much smaller proportion of the workforce. It is important to keep note of this uneven distribution when interpreting later results, as proportional comparisons across age groups may be influenced by differences in group size.

**Additional Files for Marker Review**

In addition to the main analysis report, the following files are provided for marker review.

-   README.txt: This file contains detailed documentation of the project, including data origin, files included, intended use, assumptions, limitations, and reproducibility instructions.

-   data_dictionary.csv: This file describes all variables in the final processed dataset, including variable names, data types, descriptions and derivation methods.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### **Motivation**

Generative AI has rapidly become one of the most significant technological advancements shaping modern workplaces. In Australia, AI adoption is increasing across industries, with Jobs and Skills Australia highlighting generative AI as a key force likely to influence how work is performed rather than simply replacing jobs.

However, AI exposure is not expected to be uniform across the workforce. Different occupations involve different task structures, meaning some roles are more likely to be affected by AI systems than others. At the same time, the demographic composition of occupations may shape which groups of workers are most represented in highly AI-exposed roles.

This blog explores how generative AI exposure varies across occupations in the Australian workforce, and how demographic characteristics such as age and sex are distributed across occupations with differing levels of AI exposure. It also compares two different types of AI exposure: augmentation (where AI supports human work) and automation (where AI replaces tasks).

### Data

This analysis combines two publicly available datasets.

The first is the Australian Bureau of Statistics (ABS) 2021 Census G60 Occupation by Age by Sex dataset. It provides counts of employed persons across Australian states, age groups, sex, and occupation categories. This dataset provides a detailed view of the demographic structure of the Australian workforce.

The second dataset is the Jobs and Skills Australia (JSA) Generative AI Exposure dataset. It contains occupation-level estimates of AI exposure, including measures of augmentation and automation potential. These scores are model-based estimates of how susceptible occupational tasks are to generative AI technologies.

The datasets were merged using ANZSCO major occupation group classifications to ensure comparability between occupational structures.

### Analysis

**Which occupations exhibit the highest and lowest levels of generative AI exposure scores?**

```{r}
occupation_exposure <- processed_data %>%
  distinct(major_occupation_group,
           ai_augmentation_score,
           ai_automation_score)

ggplot(
  occupation_exposure,
  aes(
    x = reorder(
      major_occupation_group,
      ai_augmentation_score
    ),
    y = ai_augmentation_score
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "AI Augmentation Exposure Score by Occupation",
    x = "Occupation Group",
    y = "AI Augmentation Exposure Score"
  )

ggplot(
  occupation_exposure,
  aes(
    x = reorder(
      major_occupation_group,
      ai_automation_score
    ),
    y = ai_automation_score
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "AI Automation Exposure Score by Occupation",
    x = "Occupation Group",
    y = "AI Automation Exposure Score"
  )
```

To examine how generative AI exposure varies across occupations, AI augmentation and AI automation exposure scores were compared across major occupation groups. This approach allows for a more nuanced understanding of how AI is likely to influence different types of work, distinguishing between AI that supports and enhances tasks (augmentation) and AI that automates or replaces tasks (automation).

The results show a clear ranking pattern across the different major occupation groups, but with important differences between augmentation and automation exposure. Professionals, Clerical and Administrative Workers, and Managers record the highest augmentation exposure scores (0.693, 0.692, and 0.691 respectively), indicating that these occupations are most likely to experience AI as a supportive tool that enhances productivity, decision-making, and knowledge-based tasks.

In contrast, automation exposure is highest for Clerical and Administrative Workers (0.600), Sales Workers (0.475), and Managers (0.395). This suggests that clerical roles in particular show higher automation exposure, reflecting the routine and structured nature of administrative tasks. Sales and managerial occupations also show moderate automation exposure, although still lower than their augmentation scores.

Occupations involving manual labour and technical work, such as Machinery Operators and Drivers (augmentation: 0.548, automation: 0.275) and Labourers (augmentation: 0.456, automation: 0.196), consistently record the lowest exposure scores across both dimensions. This indicates that these occupations are less directly affected by generative AI technologies, likely due to their reliance on physical rather than information-based tasks.

Overall, the comparison highlights a key distinction in how AI exposure manifests across the workforce. High-skilled, knowledge-intensive occupations tend to exhibit higher augmentation potential, suggesting AI will primarily complement rather than replace human work. In contrast, clerical occupations stand out as having relatively high exposure to both augmentation and automation, making them the most broadly affected group. This divergence emphasises that AI integration in the Australian workforce is not uniformly distributed and depends strongly on the structure of the occupational tasks.

**Which demographic groups (age and sex) are most concentrated in occupations with high AI-exposure?**

```{r}

high_exposure <- occupation_exposure %>%
  mutate(high_exposure = ai_augmentation_score > median(ai_augmentation_score))

demographic_exposure <- processed_data %>%
  group_by(sex, age_group, major_occupation_group) %>%
  summarise(total = sum(count), .groups = "drop") %>%
  left_join(high_exposure, by = "major_occupation_group")

demographic_totals <- processed_data %>%
  group_by(sex, age_group) %>%
  summarise(total_all = sum(count), .groups = "drop")

high_counts <- demographic_exposure %>%
  filter(high_exposure) %>%
  group_by(sex, age_group) %>%
  summarise(high_total = sum(total), .groups = "drop")

concentration <- demographic_totals %>%
  left_join(high_counts, by = c("sex", "age_group")) %>%
  mutate(
    high_total = ifelse(is.na(high_total), 0, high_total),
    proportion_high = high_total / total_all
  )

sex_concentration <- concentration %>%
  group_by(sex) %>%
  summarise(proportion_high = sum(high_total) / sum(total_all))

ggplot(sex_concentration, aes(x = sex, y = proportion_high)) +
  geom_col() +
  labs(
    title = "Sex Concentration in High AI-Exposure Occupations",
    x = "Sex",
    y = "Proportion in High Exposure Occupations"
  )

age_concentration <- concentration %>%
  group_by(age_group) %>%
  summarise(proportion_high = sum(high_total) / sum(total_all))

ggplot(age_concentration, aes(x = age_group, y = proportion_high)) +
  geom_col() +
  labs(
    title = "Age Group Concentration in High AI-Exposure Occupations",
    x = "Age Group",
    y = "Proportion in High Exposure Occupations"
  )
```

To examine how generative AI exposure is distributed across demographic groups, occupations were classified as high exposure relative to the median AI augmentation score, creating a binary grouping of occupations into above-median and below-median exposure categories.

The results show clear differences in the distribution of workers across demographic groups.

**Sex**

The findings show that female workers are more concentrated in high AI-exposure occupations compared to male workers. Approximately 69.9% of female workers are employed in above-median AI augmentation occupations, compared with 49.8% of male workers.

This suggests that women are more likely than men to be working in occupations that are more strongly associated with higher generative AI exposure scores, particularly in roles such as clerical and administrative occupations, which tend to have higher exposure scores.

**Age**

The age-based results reveal a general increasing trend in AI exposure concentration with age. Younger workers aged 15-19 (43.1%) and 20-24 (46.3%) have the lowest proportion in high-exposure occupations.

In contrast, workers in older age groups show significantly higher concentrations, with the proportion rising steadily across age bands. The highest proportions are observed among the older groups: 75-84 years (74.2%) and 85+ years (81.5%). This pattern suggests that generative AI exposure is more concentrated among older workers, likely reflecting their stronger representation in professional, managerial, and administrative occupations.

Overall, the results indicate clear demographic stratification in occupational AI exposure. However, these differences primarily reflect occupational segregation by age and sex, rather than direct demographic effects on AI exposure itself.

**To what extent are highly AI-exposed occupations characterised by AI augmentation versus AI automation?**

```{r}
ai_direction <- processed_data %>%
  group_by(major_occupation_group) %>%
  summarise(ai_balance = first(ai_balance))


ggplot(ai_direction, aes(x = reorder(major_occupation_group, ai_balance), y = ai_balance)) +
  geom_col() +
  coord_flip() + 
  labs(
    title = "Balance Between AI Augmentation and Automation Exposure by Occupation",
    x = "Major Occupation Group",
    y = "Augmentation minus Automation (AI balance)"
  )
```

Across all major occupation groups, augmentation exposure scores exceed automation exposure scores, resulting in positive AI balance values. This suggests that generative AI is currently more associated with task augmentation than task replacement in the Australian workforce.

### **Conclusion**

Overall, this analysis showed that generative AI exposure in the Australian workforce is unevenly distributed and closely associated with occupational structure. AI exposure varies systematically depending on the types of occupations in which different demographic groups are employed.

Female workers are more concentrated in high AI-exposure occupations compared to male workers, largely reflecting their higher representation in clerical and administrative roles, which tend to have higher exposure scores. Age patterns also show variation, with older workers generally having higher concentrations in high-exposure occupations. However, this pattern should be interpreted with caution, as it likely reflects the differences in occupational structure and workforce participation rather than a direct effect of age on AI exposure.

Across occupations, AI exposure is consistently more strongly characterised by augmentation than automation, suggesting that generative AI is more likely to support and enhance work tasks rather than fully replace them. Even in occupations with higher automation estimation, augmentation remains the dominant form of exposure.

These findings suggest that demographic differences in AI exposure are primarily driven by occupational distribution rather than demographic characteristics themselves. The impact of generative AI on the Australian workforce is therefore best understood as a structural issue, shaped by where different groups are employed rather than inherent differences between demographic groups themselves.

## **References**                                                                                                              

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
